Final Project Report: Developing an AI-Assisted Grading System Using Large Language Models

### 2025-12-03 FinalProject/src/paper.tex
UpdatedDrHNotes

Students directly upload \textit{free-form} assignments which are automatically tied to the uploader's account, and therefore, need no identity extraction. In these assignments, students identify crops that contain their answers. But beyond that, submissions are handled similarly to Filled-form submissions.

@@

The motivation for adoption by instructors is time-saving. If instructors do not see significant time savings, they are unlikely to adopt a new technology regardless of the other benefits. Never-the-less, additional benefits often factor into decisions once teachers realize time-saving benefits. These benefits include:

@@

\begin{compactitem}
  \item An \emph{OCR-minimal}, vision-first pipeline that uses OCR only for identity assignment.
  \item A \emph{token-budgeted batching} strategy (downscale + tiling) to bound latency/cost at class scale.
  \item A \emph{traceable data model \& APIs} with idempotent re-uploads and stable crop filenames.
  \item An \emph{instructor-in-the-loop} UX with editable groups, explicit review status, and rubric-first grading.
\end{compactitem}

@@

We target short-answer problems with recognizable visual structure (boxes/lines). Free-form essays are supported via uploaded PDFs but are not auto-scored; the system focuses on grouping to speed human grading. We assume class rosters are available, that instructors can define crops once per assignment for filled-form assignments, and that students can define crops for Free-form assignments.

The scope includes the AI grouping assistance, and workflow to grade both Filled-form and Free-form assignments.

@@

\section{Bias, Fairness, and Student Perceptions}
Bias can propagate into grading unless monitored and mitigated~\cite{Mehrabi2021}. Student acceptance depends on clear processes, and fairness~\cite{Tossell2023}. Effective formative feedback principles—timely, specific, actionable—remain central whether drafted by AI or humans~\cite{Nicol2006}.

\section{Similar Implementations}
To situate this project, we survey adjacent tools and how our system compares. In short,
\emph{we have not found public documentation of a production system that groups handwritten
short answers directly from images using a general-purpose vision LLM without first using OCR on the
answer content}. Existing offerings fall into four families:

@@

\section{High-Level Architecture}
At a glance, the platform is organized into three dashed \emph{zones} that separate concerns and scaling boundaries: the \emph{Application Server} (Razor Pages web app and extraction worker), the \emph{AI Service Layer} (FastAPI microservice and the GPT-4o Vision endpoint), and \emph{Data \& Storage} (MySQL and the file store). Instructors interact via a browser over HTTPS with the Razor Pages app, which handles authentication, rubric management, and the grading UI. The app invokes an extraction worker that orchestrates Ghostscript, PdfPig, and SkiaSharp to render pages and cut out per-answer crops; Tesseract OCR is used narrowly to read identifying fields (e.g., student ID) and is deliberately decoupled from semantic grouping. Crops and downstream exports are written to the file store while the web app serves these images directly to the UI.

@@

\section{Token Budget, Cost \& Rate Limiting}
For an image of width $w$ and height $h$, the service estimates tokens after a 50\% downscale: $w' = \lfloor w/2 \rfloor$, $h' = \lfloor h/2 \rfloor$. The number of 512$\times$512 tiles is $T=\lceil w'/512 \rceil\cdot\lceil h'/512 \rceil$, and the cost estimate is $85 + 170\,T$ tokens/image. Batches exceeding a threshold are split. On rate limits, the client retries up to 10 times with exponential backoff. For a representative $768\times 768$ downscaled image ($T=3$), the estimate is $\sim595$ tokens/image.

@@

\section{Objectives}
We provide concise, visual evidence that the system:
\begin{enumerate}
  \item processes submissions end-to-end,
  \item groups answer \emph{images} by semantic similarity (no answer OCR\footnote{No answer OCR means that text extracted from the image is not returned in the answer. In this case the answer is a grouping, so text extracted by the AI is not needed.}),
  \item isolates outliers for review,
  \item applies rubrics and propagates score changes consistently, and
  \item supports clipboard-based grade transfer (\emph{Copy Table}) into external systems.
\end{enumerate}

@@

\section{Clipboard Interoperability (Evidence)}

@@

\paragraph{\emph{On variability of student answers.}}
Cluster counts naturally vary with how diverse student responses are for a given question (e.g., multiple valid solution paths or many non-answers increase the number of clusters/ungrouped). \emph{Even when clusters proliferate, presenting all answers for one question on a single screen still reduces handling time versus grading one student at a time or flipping through pages/PDFs}, because instructors can scan, compare, and bulk-apply rubric actions across visually similar answers.

@@

\section{Final Outcomes and Deliverables}
We delivered the integrated web app, background services, reproducible prompts/model versions, and a \emph{Copy Table} flow that enables fast pasting of grades into external systems (e.g., LMS/Sheets). Documentation includes an instructor guide and technical deployment notes.

\section{Lessons Learned and Future Work}
\begin{compactitem}
  \item Visual pre-processing (downscale/tiling) mattered more for stability than minor prompt tuning.
  \item Instructors preferred neutral group names and an explicit \emph{Ungrouped} bin.
  \item Future: direct CSV/Excel export in addition to \emph{Copy Table}; domain-tuned prompts for math diagrams; adaptive thresholding for faint pencil; pre-clustering to cut LLM calls; regrade workflow.
  \item Future: prompting with answers to identify the correct group.
  \item Future: prompting with rubrics that the AI can assign to individual answers or groups. 
\end{compactitem}

>>>>

Students directly upload free-form assignments which are automatically tied to the uploader's account and therefore need no identity extraction. In these assignments, students identify crops that contain their answers. But beyond that, submissions are handled similarly to filled-form submissions.

@@

The motivation for adoption by instructors is time-saving. If instructors do not see significant time savings, they are unlikely to adopt a new technology regardless of the other benefits. Nevertheless, additional benefits often factor into decisions once teachers realize time-saving benefits. These benefits include:

@@

\begin{compactitem}
  \item An \emph{OCR-minimal}, vision-first pipeline that uses OCR only for identity assignment.
  \item A \emph{token-budgeted batching} strategy (downscale + tiling) to bound latency/cost at class scale.
  \item A \emph{traceable data model and APIs} with idempotent re-uploads and stable crop filenames.
  \item An \emph{instructor-in-the-loop} UX with editable groups, explicit review status, and rubric-first grading.
\end{compactitem}

@@

We target short-answer problems with recognizable visual structure (boxes/lines). Free-form essays are supported via uploaded PDFs but are not auto-scored; the system focuses on grouping to speed human grading. We assume class rosters are available, that instructors can define crops once per assignment for filled-form assignments, and that students can define crops for free-form assignments.

The scope includes the AI grouping assistance, and workflow to grade both filled-form and free-form assignments.

@@

\section{Bias, Fairness, and Student Perceptions}
Bias can propagate into grading unless monitored and mitigated~\cite{Mehrabi2021}. Student acceptance depends on clear processes and fairness~\cite{Tossell2023}. Effective formative feedback principles—timely, specific, actionable—remain central whether drafted by AI or humans~\cite{Nicol2006}.

\section{Similar Implementations}
To situate this project, we survey adjacent tools and how our system compares. In short, we have not found public documentation of a production system that groups handwritten
short answers directly from images using a general-purpose vision LLM without first using OCR on the
answer content. Existing offerings fall into four families:

@@

\section{High-Level Architecture}
At a glance, the platform is organized into three dashed \emph{zones} that separate concerns and scaling boundaries (Figure~\ref{fig:architecture}): the \emph{Application Server} (Razor Pages web app and extraction worker), the \emph{AI Service Layer} (FastAPI microservice and the GPT-4o Vision endpoint), and \emph{Data \& Storage} (MySQL and the file store). Instructors interact via a browser over HTTPS with the Razor Pages app, which handles authentication, rubric management, and the grading UI. The app invokes an extraction worker that orchestrates Ghostscript, PdfPig, and SkiaSharp to render pages and cut out per-answer crops; Tesseract OCR is used narrowly to read identifying fields (e.g., student ID) and is deliberately decoupled from semantic grouping. Crops and downstream exports are written to the file store while the web app serves these images directly to the UI.

@@

\section{Token Budget, Cost \& Rate Limiting}
For an image of width $w$ and height $h$, the service estimates tokens after a 50\% downscale: $w' = \lfloor w/2 \rfloor$, $h' = \lfloor h/2 \rfloor$. The number of 512$\times$512 tiles is $T=\lceil w'/512 \rceil\cdot\lceil h'/512 \rceil$, and the cost estimate is $85 + 170\,T$ tokens/image. Batches exceeding a threshold are split. On rate limits, the client retries up to 10 times with exponential backoff. For a representative $768\times 768$ downscaled image ($T=3$), the estimate is approximately 595 tokens/image.

@@

\section{Objectives}
We provide concise, visual evidence that the system:
\begin{enumerate}
  \item processes submissions end-to-end,
  \item groups answer images by semantic similarity (no answer OCR\footnote{No answer OCR means that text extracted from the image is not returned in the answer. In this case the answer is a grouping, so text extracted by the AI is not needed.}),
  \item isolates outliers for review,
  \item applies rubrics and propagates score changes consistently, and
  \item supports clipboard-based grade transfer (\emph{Copy Table}) into external systems.
\end{enumerate}

@@

@@

\paragraph{\emph{On variability of student answers.}}
Cluster counts naturally vary with how diverse student responses are for a given question (e.g., multiple valid solution paths or many non-answers increase the number of clusters/ungrouped). Even when clusters proliferate, presenting all answers for one question on a single screen still reduces handling time versus grading one student at a time or flipping through pages/PDFs, because instructors can scan, compare, and bulk-apply rubric actions across visually similar answers.

@@

\section{Final Outcomes and Deliverables}
We delivered the integrated web app, background services, a reproducible AI configuration (prompts and model settings), and a \emph{Copy Table} flow that enables fast pasting of grades into external systems (e.g., LMSs or Sheets). Documentation includes an instructor guide and technical deployment notes.

\section{Lessons Learned and Future Work}
Implementing and using the system with instructors surfaced several practical insights about what most affected robustness and instructor experience, as well as concrete opportunities to extend the platform. Below we summarize key lessons and outline directions for future work.

\begin{compactitem}
  \item Visual pre-processing (downscale/tiling) mattered more for stability than minor prompt tuning.
  \item Instructors preferred neutral group names and an explicit \emph{Ungrouped} bin.
  \item Future: direct CSV/Excel export in addition to \emph{Copy Table}; domain-tuned prompts for math diagrams; adaptive thresholding for faint pencil; pre-clustering to cut LLM calls; regrade workflow.
  \item Future: prompting with answers to identify the correct group.
  \item Future: prompting with rubrics that the AI can assign to individual answers or groups. 
\end{compactitem}
